A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-20 sections/Overview.tex
declarative typing description draft

The syntax of this declarative system is shown in Figure~\ref{fig:decl:syntax}.
A declarative type $A$ is either the unit type $1$, a type variable $a$,
a universal quantification $\forall a. A$ or a function type $A \to B$.

Nested universal quantifiers are allowed for types,
but monotypes $\tau$ do not have any universal quantifier.
Terms include a unit term $()$, variables $x$, lambda-functions $\lam e$,
applications $e_1~e_2$ and annotations $(e:A)$.
Contexts $\Psi$ are sequences of type variable declarations and
term variables with its type declared $x:A$.

@@

\begin{figure}[t]
\centering \framebox{$\Psi \vdash A \Lto B$} $e$ checks against input type $A$.\\
\centering \framebox{$\Psi \vdash A \To B$} $e$ synthesizes output type $A$.\\
\centering \framebox{$\Psi \vdash \appInf{A}{e}{C}$} Applying a function of type $A$ to $e$ synthesizes type $C$.
\begin{gather*}
\inferrule*[right=$\mathtt{DeclVar}$]
    {(x:A)\in\Psi}{\Psi\vdash x\To A}
\qquad
\inferrule*[right=$\mathtt{DeclSub}$]
%e \neq \lam e' \quad B \neq \all B' \quad 
    {\Psi\vdash e\To A \quad \Psi\vdash A\le B}
    {\Psi \vdash e\Lto B}
\\
\inferrule*[right=$\mathtt{DeclAnno}$]
    {\Psi \vdash A \quad \Psi\vdash e\Lto A}
    {\Psi\vdash (e:A)\To A}
\qquad
\inferrule*[right=$\mathtt{Decl1I}$]
    {~}{\Psi\vdash () \Lto 1}
\qquad
\inferrule*[right=$\mathtt{Decl1I\To}$]
    {~}{\Psi\vdash () \To 1}
\\
\inferrule*[right=$\mathtt{Decl\forall I}$]
    {\Psi,a \vdash e \Lto A}
    {\Psi\vdash e\Lto \all A}
\qquad
\inferrule*[right=$\mathtt{Decl\forall App}$]
    {\Psi \vdash \tau \quad \Psi\vdash \appInf{[\tau/a]A}{e}{C} }
    {\Psi\vdash \appInf{\all A}{e}{C}}
\\
\inferrule*[right=$\mathtt{Decl\to I}$]
    {\Psi,x:A \vdash e\Lto B}
    {\Psi\vdash \lam e \Lto A \to B}
\qquad
\inferrule*[right=$\mathtt{Decl\to I\To}$]
    {\Psi\vdash \sigma\to\tau \quad \Psi,x:\sigma \vdash e\Lto \tau}
    {\Psi\vdash \lam e \To \sigma\to\tau}
\\
\inferrule*[right=$\mathtt{Decl\to E}$]
    {\Psi\vdash e_1\To A \quad \Psi\vdash \appInf{A}{e_2}{C}}
    {\Psi\vdash e_1~e_2 \To C}
\qquad
\inferrule*[right=$\mathtt{Decl\to App}$]
    {\Psi\vdash e \Lto A}
    {\Psi\vdash \appInf{A \to C}{e}{C}}
\end{gather*}
\caption{Declarative Typing}\label{fig:decl:typing}
\end{figure}

@@

\paragraph{Declarative Typing}
\jimmy{TODO decl. typing description}

\paragraph{Overlapping Rules}
\jimmy{Move to the Alg. Section}
Some of the declarative rules overlaps with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \all A \le \all B$.
In such case, choose $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.

Declarative typing rule $\mathtt{DeclSub}$ overlaps with
both $\mathtt{Decl\forall I}$ and $\mathtt{Decl\to I}$.
However, we could argue that more specific rules are always the best choices,
i.e. $\mathtt{Decl\forall I}$ and $\mathtt{Decl\to I}$ should have
higher priority to $\mathtt{DeclSub}$.
For example, $\Psi\vdash \lam x \Lto \all a\to a$ succeeds if derived from
Rule $\mathtt{Decl\forall I}$, but fails when applied to $\mathtt{DeclSub}$:
$$
\inferrule*[right={$\mathtt{Decl\forall I}$}]
	{\inferrule*[Right={$\mathtt{Decl\to I}$}]
		{\Psi,a,x:a \vdash x \Lto a}
		{\Psi,a \vdash \lam x \Lto a \to a}
	}
	{\Psi\vdash \lam x \Lto \forall a \to a}
$$
$$
\inferrule*[right={$\mathtt{DeclSub}$}]
	{
		\inferrule*[right=$\mathtt{Decl\to I\To}$]
			{\Psi \vdash \blue\sigma\to \blue\tau \quad \Psi,x:\blue\sigma\vdash e \Lto \blue\tau}
			{\Psi \vdash \lam x \To \blue\sigma\to \blue\tau}\quad
		\inferrule*[Right=$\mathtt{{\le}\forall R}$]
			{\inferrule*[Right=$\mathtt{{\le}{\to}}$]
				{
					\inferrule*[Right=$?$]
						{\text{Impossible! }\blue\sigma \neq a}
						{\Psi,a \vdash a \le \blue\sigma}
					\quad \Psi,a \vdash \blue\tau \le a
				}
				{\Psi,a \vdash \blue\sigma\to \blue\tau \le a \to a}
			}
			{\Psi\vdash \blue\sigma\to \blue\tau\le \all a \to a}
	}
{\Psi\vdash \lam x \Lto \forall a \to a}
$$

>>>>

The syntax of this declarative system is shown in Figure~\ref{fig:decl:syntax}.
A declarative type $A$ is either the unit type $1$, a type variable $a$,
a universal quantification $\forall a. A$ or a function type $A \to B$.
Nested universal quantifiers are allowed for types,
but monotypes $\tau$ do not have any universal quantifier.
Terms include a unit term $()$, variables $x$, lambda-functions $\lam e$,
applications $e_1~e_2$ and annotations $(e:A)$.
Contexts $\Psi$ are sequences of type variable declarations and
term variables with its type declared $x:A$.

@@

\begin{figure}[t]
\begin{tabular}{rl}
    \framebox{$\Psi \vdash A \Lto B$} & $e$ checks against input type $A$.\\[0.5mm]
    \framebox{$\Psi \vdash A \To B$} & $e$ synthesizes output type $A$.\\[0.5mm]
    \framebox{$\Psi \vdash \appInf{A}{e}{C}$} & Applying a function of type $A$ to $e$ synthesizes type $C$.
\end{tabular}
\begin{gather*}
\inferrule*[right=$\mathtt{DeclVar}$]
    {(x:A)\in\Psi}{\Psi\vdash x\To A}
\qquad
\inferrule*[right=$\mathtt{DeclSub}$]
%e \neq \lam e' \quad B \neq \all B' \quad 
    {\Psi\vdash e\To A \quad \Psi\vdash A\le B}
    {\Psi \vdash e\Lto B}
\\
\inferrule*[right=$\mathtt{DeclAnno}$]
    {\Psi \vdash A \quad \Psi\vdash e\Lto A}
    {\Psi\vdash (e:A)\To A}
\qquad
\inferrule*[right=$\mathtt{Decl1I}$]
    {~}{\Psi\vdash () \Lto 1}
\qquad
\inferrule*[right=$\mathtt{Decl1I\To}$]
    {~}{\Psi\vdash () \To 1}
\\
\inferrule*[right=$\mathtt{Decl\forall I}$]
    {\Psi,a \vdash e \Lto A}
    {\Psi\vdash e\Lto \all A}
\qquad
\inferrule*[right=$\mathtt{Decl\forall App}$]
    {\Psi \vdash \tau \quad \Psi\vdash \appInf{[\tau/a]A}{e}{C} }
    {\Psi\vdash \appInf{\all A}{e}{C}}
\\
\inferrule*[right=$\mathtt{Decl\to I}$]
    {\Psi,x:A \vdash e\Lto B}
    {\Psi\vdash \lam e \Lto A \to B}
\qquad
\inferrule*[right=$\mathtt{Decl\to I\To}$]
    {\Psi\vdash \sigma\to\tau \quad \Psi,x:\sigma \vdash e\Lto \tau}
    {\Psi\vdash \lam e \To \sigma\to\tau}
\\
\inferrule*[right=$\mathtt{Decl\to E}$]
    {\Psi\vdash e_1\To A \quad \Psi\vdash \appInf{A}{e_2}{C}}
    {\Psi\vdash e_1~e_2 \To C}
\qquad
\inferrule*[right=$\mathtt{Decl\to App}$]
    {\Psi\vdash e \Lto A}
    {\Psi\vdash \appInf{A \to C}{e}{C}}
\end{gather*}
\caption{Declarative Typing}\label{fig:decl:typing}
\end{figure}

@@

\paragraph{Declarative Typing}
The bidirectional type system has three judgments shown in Figure~\ref{fig:decl:typing}.
The checking judgment $\Psi\vdash e\Lto A$ checks $e$ against the type $A$ under the context $\Psi$.
The synthesis judgment $\Psi\vdash e\To A$ means that we can infer the type of $e$ under the context $\Psi$.
The application judgment $\Psi\vdash \appInf{A}{e}{C}$ means that if a function of type $A$
(which could be a polymorphic one) is applied to the argument $e$, then the whole application
synthesizes type $C$ as its return type.

Some of the rules handles simple cases with common approaches.
Rule $\mathtt{DeclVar}$ looks term variables up in the context.
Rules $\mathtt{Decl1I}$ and $\mathtt{Decl1I{\To}}$ checks and synthesizes the unit type, respectively.
Rule $\mathtt{DeclAnno}$ synthesizes the type of the annotated expression $(e:A)$
to its annotated type $A$, and requires $e$ to be checked agains $A$.
Checking an expression $e$ against a polymorphic type $\all A$ under the context $\Psi$ succeeds
if $e$ checks against A under an extended context $(\Psi, a)$, where $a$ is picked as a free variable.

The subsumption rule $\mathtt{DeclSub}$ depends on the subtyping relation,
and changes mode from checking to synthesis: if $e$ synthesizes $A$ and $A\le B$
($A$ is more polymorphic than $B$), then $e$ can be checked against $B$.
If a checking problem does not match any other rules,
$\mathtt{DeclSub}$ can be applied and try to synthesize a type
before checking whether the synthesized one is more general.

Lambda-functions is a critical component of bidirectional type system, and they require much attention.
Checking $\lam e$ against function type $A\to B$ is simple:
we check the body $e$ against $B$ under the context extended with $x:A$.
However, synthesizing a lambda-function is not easy in general,
and this \emph{predicative} declarative system is able to
synthesize to a suitable monotype $\sigma\to\tau$.
In other words, if no type information is given for a lambda function,
we will synthesize it to the best mono-typed function according to the context,
which also requires some guessing similar to the subtyping rule $\mathtt{\forall L}$.

Application $(e_1~e_2)$ is handled by Rule $\mathtt{Decl{\to}E}$,
which firstly synthesize type $A$ for the function $e_1$.
If $A$ is a function type $B\to C$, Rule $\mathtt{Decl{\to}App}$ is applied,
by checking the argument $e_2$ against $B$ and returns type $C$.
The only other possibile kind of types that
the function $e_1$ could be synthesized is polymorphic types $\all A$.
In that case, we instantiate $A$ to $[\tau/a]A$ with a monotype $\tau$ (which is also guessed)
according to Rule $\mathtt{Decl{\to}I{\To}}$.
If $[\tau/a]A$ is a function type, Rule $\mathtt{Decl{\to}App}$ proceeds;
if $[\tau/a]A$ is still polymorphic, Rule $\mathtt{Decl{\to}I{\To}}$ is recursively applied.

\jimmy{Possibly some discussions about the typing rules}

\paragraph{Overlapping Rules}
\jimmy{Mention the resolution of overlapping briefly, and then point to algorithm and metatheory}
Some of the declarative rules overlaps with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \all A \le \all B$.
In such case, choose $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.

Declarative typing rule $\mathtt{DeclSub}$ overlaps with
both $\mathtt{Decl\forall I}$ and $\mathtt{Decl\to I}$.
However, we could argue that more specific rules are always the best choices,
i.e. $\mathtt{Decl\forall I}$ and $\mathtt{Decl\to I}$ should have
higher priority to $\mathtt{DeclSub}$.
For example, $\Psi\vdash \lam x \Lto \all a\to a$ succeeds if derived from
Rule $\mathtt{Decl\forall I}$, but fails when applied to $\mathtt{DeclSub}$:
$$
\inferrule*[right={$\mathtt{Decl\forall I}$}]
	{\inferrule*[Right={$\mathtt{Decl\to I}$}]
		{\Psi,a,x:a \vdash x \Lto a}
		{\Psi,a \vdash \lam x \Lto a \to a}
	}
	{\Psi\vdash \lam x \Lto \all a \to a}
$$
$$
\inferrule*[right={$\mathtt{DeclSub}$}]
	{
		\inferrule*[right=$\mathtt{Decl\to I\To}$]
			{\Psi \vdash \blue\sigma\to \blue\tau \quad \Psi,x:\blue\sigma\vdash e \Lto \blue\tau}
			{\Psi \vdash \lam x \To \blue\sigma\to \blue\tau}\quad
		\inferrule*[Right=$\mathtt{{\le}\forall R}$]
			{\inferrule*[Right=$\mathtt{{\le}{\to}}$]
				{
					\inferrule*[Right=$?$]
						{\text{Impossible! }\blue\sigma \neq a}
						{\Psi,a \vdash a \le \blue\sigma}
					\quad \Psi,a \vdash \blue\tau \le a
				}
				{\Psi,a \vdash \blue\sigma\to \blue\tau \le a \to a}
			}
			{\Psi\vdash \blue\sigma\to \blue\tau\le \all a \to a}
	}
{\Psi\vdash \lam x \Lto \all a \to a}
$$
